Design Document: Monte Carlo Maximum Likelihood for Generalized Linear Mixed Models

### 2015-01-23 NewDesignDoc/DesignDocGLMM.tex
Added constants in to f-twiddle

 We can  express the log unnormalized pdf of $N(0, D)$  as:
\begin{align}
\log f_\theta(u)= \log f (u| 0, D) = (-1/2) \log |D| - (1/2) u' D^{-1} u
\end{align}

@@

\subsection{Evaluating the importance sampling distribution}
This section discusses evaluating the importance sampling distribution shown in \ref{eq:ftwiddle}. Section \ref{sec:Dgeneral} describes how to evaluate the pdf for the normal components of $\tilde{f}(u)$. Section \ref{sec:tstuff} describes how to evaluate the pdf for the t component of $\tilde{f}(u)$. 
The  $\tilde{f}(u)$ calculation is used in equation \ref{eq:MCLA}.

%%distRandGeneral
\subsubsection{Normal distribution for general variance matrix $\Sigma$ (\texttt{distRandGeneral})}\label{sec:Dgeneral}
Let $\Sigma$ be a variance matrix and let $\mu$ be a mean vector. Consider $N(\mu, \Sigma)$.   We can  write the log pdf of this distribution  as:
\begin{align}
\log f (u| \mu, \Sigma) = (1/2) \log |\Sigma^{-1}| - (1/2) (u-\mu)' \Sigma^{-1} (u-\mu)
\end{align}
 The only part of this to discuss is $|\Sigma^{-1}|$. We can use eigendecomposition to make $\Sigma^{-1}=O \Lambda O'$ where $O$ is orthogonal and $\Lambda$ is the diagonal matrix with eigenvalues. Since orthogonal matrices have determinant $\pm 1$, then $|O||O'|=1$. Thus  
\begin{align}
|\Sigma^{-1}|&=|O'| \; |\Lambda| \; |O| \\
&= |O'| \; |O| \; |\Lambda| \\
&=|\Lambda|,
\end{align}
which is just the product of the eigenvalues. The log of the determinant is calculated beforehand to save time, since this function is called $3m$ times throughout each trust optimization iteration, where $m$ is again the Monte Carlo sample size.

@@

Consider $t_\zeta$. We can write the log unnormalized pdf of this distribution as:
\begin{align}
\log \grave{f}(u)&=\log \left[ 1+u'D^* u/\zeta  \right]^{-(\zeta+q)/2}\\
&=-\left[ \dfrac{\zeta+q}{2} \right] \log (1+u'D^*u/\zeta)\\
&= -\left[ \dfrac{\zeta+q}{2} \right] \log \left(1+\sum_{t=1}^T \dfrac{u_t'u_t}{\zeta \nu_t}\right)
\end{align}

Rewritten in C, we need to pass in \texttt{u} as a pointer, \texttt{zeta} as an int, \texttt{myq} as an int array, \texttt{meow} as an int array, and \texttt{logft} as a pointer.

@@

&\leq \int \dfrac{f_\theta(y|u)^2 f_\theta(u)^2}{p_1 \grave{f}(u)} du\\

@@

First, we start with \eqref{eq:first}:
\begin{align}
\int \dfrac{ \left[ \nabla \log f_\theta (y|u) \right]^2 \left[ f_\theta(u,y)\right]^2}{\tilde{f}(u)} du 
&=\int \dfrac{ \left[ \nabla \log f_\theta (y|u) \right]^2 \left[ f_\theta(y|u)\right]^2 \left[ f_\theta(u)\right]^2}{\tilde{f}(u)} du \\
&\leq \int \dfrac{ \left[ \nabla \log f_\theta (y|u) \right]^2 \left[ f_\theta(y|u)\right]^2 \left[ f_\theta(u)\right]^2}{p_1\grave{f}(u)} du \\
&\propto \int \dfrac{ \left[ \nabla \log f_\theta (y|u) \right]^2 \left[ f_\theta(y|u)\right]^2 \exp(-u'D^{-1}u)}{p_1\left[ 1+u'D^*u/\zeta  \right]^{-(\zeta+q)/2}} du \\
&= \int \dfrac{ \left[ \nabla \log f_\theta (y|u) \right]^2 \left[ f_\theta(y|u)\right]^2 \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} }{p_1 \exp(u'D^{-1}u)} du \\
&= \int \dfrac{ \left[ \nabla \log f_\theta (y|u) \right]^2 \, \exp(2y'(X \beta+Zu)) \,  \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} }{ p_1 \, \exp(2 c(\eta)) \,\exp(u'D^{-1}u)} du \label{eq:soon}
\end{align}
By the Cauchy-Schwartz inequality, the integral in \eqref{eq:soon} exists if
\begin{align}
 & \int \dfrac{ \left[\dfrac{\partial}{\partial \beta} \log f_\theta (y|u) \right]^2 \, \exp(2y'(X \beta+Zu)) \,  \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} }{ p_1 \, \exp(2 c(\eta)) \,\exp(u'D^{-1}u)} du \nonumber \\
&\propto  \int \dfrac{ \left[X' c'(\eta) \right]^2 \, \exp(2y'(X \beta+Zu)) \,  \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} }{ p_1 \, \exp(2 c(\eta)) \,\exp(u'D^{-1}u)} du \label{eq:soon2}
\end{align}
exists, where $c'(\eta)$ is a vector with components $c'(\eta_i)$.  This will converge as long as no term in the numerator of the integrand grows more quickly than $\exp(u'D^{-1}u$. The second and third terms grow less quickly than $\exp(u'D^{-1}u$, so $\left[ c'(\eta)\right]^2$ is the only term to check.  If $y|u \sim$ Bernoulli, then 
\begin{align}
c'(\eta_i)= \dfrac{\exp(\eta_i)}{1+\exp(\eta_i)},
\end{align}
which does not grow more quickly than $\exp(u'D^{-1}u)$. If $y|u \sim$ Poisson, then
\begin{align}
c'(\eta_i)= e^{\eta_i},
\end{align}
which does not grow more quickly than $\exp(u'D^{-1}u)$. Therefore, \eqref{eq:first} exists and is finite, and 
\begin{align}
\var \left( \sum_{k=1}^m \nabla \log f_\theta (y|u_k) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) < \infty.
\end{align}
is finite.

Next, we move on to showing the existence of \eqref{eq:second}.
\begin{align}
\int \dfrac{ \left[ \nabla \log f_\theta (u) \right]^2 \left[ f_\theta(u,y)\right]^2}{\tilde{f}(u)} du &= 
\int \dfrac{ \left[ \nabla \log f_\theta (u) \right]^2 \left[ f_\theta(y|u)\right]^2  \left[ f_\theta(u)\right]^2}{\tilde{f}(u)} du\\
&\leq \int \dfrac{ \left[ \nabla \log f_\theta (u) \right]^2 \left[ f_\theta(y|u)\right]^2  \left[ f_\theta(u)\right]^2}{p_1 \grave{f}(u)} du
\end{align}
By the Cauchy-Schwartz inequality, the above integral exists as long as  
\begin{align}
\int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t} \log f_\theta (u) \right]^2 \left[ f_\theta(y|u)\right]^2  \left[ f_\theta(u)\right]^2}{p_1 \grave{f}(u)} du < \infty
\end{align}
for every $t=1,\ldots, T$. We see
\begin{align}
&\int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t} \log f_\theta (u) \right]^2 \left[ f_\theta(y|u)\right]^2  \left[ f_\theta(u)\right]^2}{p_1 \grave{f}(u)} du \\
&= \int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t} \log f_\theta (u) \right]^2 \exp(2y'(X\beta+Zu)   )  \left[ f_\theta(u)\right]^2}{p_1 \exp(2c(\eta))    \grave{f}(u)} du \\
&\propto \int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t} \log f_\theta (u) \right]^2 \exp(2y'(X\beta+Zu)   ) \,   \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} }{p_1 \exp(2c(\eta)) \, \exp(u'D^{-1}u)  } du\\
&= \int \dfrac{ \left[  -\dfrac{q_t}{2 \nu_t} + \dfrac{u_t'u_t}{2 \nu_t^2}   \right]^2 \exp(2y'(X\beta+Zu)   ) \,   \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} }{p_1 \exp(2c(\eta)) \, \exp(u'D^{-1}u)  } du
\end{align}
Because every term in the numerator of the integrand grows less quickly than $\exp(u'D^{-1}u)$, the integral exists.  Therefore, \eqref{eq:second} exists and 
\begin{align}
\var \left(\sum_{k=1}^m \nabla \log f_\theta (u_k) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) < \infty.
\end{align}

Lastly, we want to show the existence of \eqref{eq:third}. We see
\begin{align}
\int \dfrac{  \nabla \log f_\theta (u)\,  \nabla \log f_\theta (y|u) \, \left[ f_\theta(u,y)\right]^2}{\tilde{f}(u)} du &=
\int \dfrac{  \nabla \log f_\theta (u) \, \nabla \log f_\theta (y|u) \, \left[ f_\theta(y|u)\right]^2 \left[ f_\theta(u)\right]^2}{\tilde{f}(u)} du \\
&\leq
\int \dfrac{  \nabla \log f_\theta (u) \,  \nabla \log f_\theta (y|u)\,  \left[ f_\theta(y|u)\right]^2 \left[ f_\theta(u)\right]^2}{p_1 \grave{f}(u)} du \label{eq:this1}
\end{align}
Again, we use the Cauchy-Schwartz inequality to check the existence of \eqref{eq:this1}. If, for every $t=1,\ldots,T$,
\begin{align}
\int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t}  \log f_\theta (u) \right] \left[  \dfrac{\partial}{\partial \beta} \log f_\theta (y|u) \right]   \left[ f_\theta(y|u)\right]^2 \left[ f_\theta(u)\right]^2}{p_1 \grave{f}(u)} du \label{eq:this1}
\end{align}
exists, then \eqref{eq:this1} also exists. Continuing, we see
\begin{align}
& \int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t}  \log f_\theta (u) \right] \left[  \dfrac{\partial}{\partial \beta} \log f_\theta (y|u) \right]   \left[ f_\theta(y|u)\right]^2 \left[ f_\theta(u)\right]^2}{p_1 \grave{f}(u)} du\\
 &\propto \int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t}  \log f_\theta (u) \right] \left[  \dfrac{\partial}{\partial \beta} \log f_\theta (y|u) \right]   \left[ f_\theta(y|u)\right]^2 }{p_1 \, \exp(u'D^{-1}u) \, \grave{f}(u)} du\\
 &\propto \int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t}  \log f_\theta (u) \right] \left[  \dfrac{\partial}{\partial \beta} \log f_\theta (y|u) \right]   \left[ f_\theta(y|u)\right]^2 \, \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} }{p_1 \, \exp(u'D^{-1}u) } du\\
&\propto \int \dfrac{ \left[      -\dfrac{q_t}{2 \nu_t} + \dfrac{u_t'u_t}{2 \nu_t^2}        \right] \left[  X'(y-c'(\eta)) \right]   \left[ f_\theta(y|u)\right]^2 \, \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} }{p_1 \, \exp(u'D^{-1}u) } du\\
&= \int \dfrac{ \left[      -\dfrac{q_t}{2 \nu_t} + \dfrac{u_t'u_t}{2 \nu_t^2}        \right] \left[  X'(y-c'(\eta)) \right]   \, \exp(2y'(X\beta+Zu)) \, \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} }{p_1 \, \exp(u'D^{-1}u) \, \exp(2 c(\eta)) } du. \label{eq:this2}
\end{align}
We have already shown that $-c'(\eta)$ does not grow more quickly than $\exp(u'D^{-1}u)$. The other terms in the numerator of the integrand shown in \eqref{eq:this2} also grow more slowly than $\exp(u'D^{-1}u)$. We conclude that \eqref{eq:this2} exists, which implies that \eqref{eq:third} exists and
\begin{align}
\cov \left(\sum_{k=1}^m \nabla \log f_\theta (y|u_k) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \, , \, \sum_{k=1}^m \nabla \log f_\theta (u_k) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) < \infty.
\end{align}
This shows that
\begin{align}
\var \left( \sum_{k=1}^m     \nabla \log f_\theta(u_k,y)\dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) < \infty. 
\end{align}
Therefore, the gradient of the MCLA has finite variance and a central limit theorem.

>>>>

 We can  express the log unnormalized pdf of $N(0, D)$  as:
\begin{align}
\log f_\theta(u)= \log f (u| 0, D) = - q/2 \log (2 \pi) -1/2 \log |D| - (1/2) u' D^{-1} u
\end{align}

@@

\subsection{Evaluating the importance sampling distribution}
This section discusses evaluating the importance sampling distribution shown in \ref{eq:ftwiddle}. The  $\tilde{f}(u)$ calculation is used in equation \ref{eq:MCLA}.  Section \ref{sec:Dgeneral} describes how to evaluate the pdf for the normal components of $\tilde{f}(u)$. Section \ref{sec:tstuff} describes how to evaluate the pdf for the t component of $\tilde{f}(u)$.

To be computationally stable, we need to be careful  with the addition of the terms in \ref{eq:ftwiddle} to prevent overflow. We know that
\begin{align}
 \tilde{f}(u) &= p_1  \grave{f}(u|0,D^*)+p_2  f(u \, | \, u^*, D^*)+p_3  f(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1}) \\
&= p_1 \exp( \log  \grave{f}(u|0,D^*) ))+p_2 \exp( \log(  f(u \, | \, u^*, D^*) ))+ \nonumber\\
&+ p_3 \exp( \log( f(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1}) )).
\end{align}
Set
\begin{align}
\tilde{a}= \max \left\{ \log  \grave{f}(u|0,D^*) ), \log  f(u \, | \, u^*, D^*) ), \log f(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})  \right\}
\end{align}
Then
\begin{align}
 \tilde{f}(u) 
&= p_1 \exp\left( \log  \grave{f}(u|0,D^*)  -\tilde{a}\right)+p_2 \exp \left( \log  f(u \, | \, u^*, D^* )-\tilde{a} \right)+ \nonumber\\
&+ p_3 \exp\left[ \log f \left(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1} \right)  -\tilde{a} \right].
\end{align}
is more stable computationally.

%%distRandGeneral
\subsubsection{Normal distribution for general variance matrix $\Sigma$ (\texttt{distRandGeneral})}\label{sec:Dgeneral}
Let $\Sigma$ be a variance matrix and let $\mu$ be a mean vector. Consider $N(\mu, \Sigma)$.   We can  write the log pdf of this distribution  as:
\begin{align}
\log f (u| \mu, \Sigma) = -\dfrac{q}{2} \log (2 \pi) + \dfrac{1}{2} \log |\Sigma^{-1}| - \dfrac{1}{2} (u-\mu)' \Sigma^{-1} (u-\mu)
\end{align}
 The only part of this to discuss is $|\Sigma^{-1}|$. We can use eigendecomposition to make $\Sigma^{-1}=O \Lambda O'$ where $O$ is orthogonal and $\Lambda$ is the diagonal matrix with eigenvalues. Since orthogonal matrices have determinant $\pm 1$, then $|O||O'|=1$. Thus  
\begin{align}
|\Sigma^{-1}|&=|O'| \; |\Lambda| \; |O| \\
&= |O'| \; |O| \; |\Lambda| \\
&=|\Lambda|,
\end{align}
which is just the product of the eigenvalues. The log of the determinant is calculated beforehand to save time, since this function is called $3m$ times throughout each trust optimization iteration, where $m$ is again the Monte Carlo sample size.

@@

Consider $t_\zeta$. We can write the log unnormalized pdf of this distribution as:
\begin{align}
\log \grave{f}(u|0,D^*)&=  \log \left( \Gamma \left( \dfrac{\zeta+q}{2}  \right)   \right)-\log \left( \Gamma \left( \dfrac{\zeta}{2}  \right)   \right)  - \dfrac{q}{2} \log(\nu \pi)\\
&- \dfrac{1}{2} \log |D^*|     -\left[ \dfrac{\zeta+q}{2} \right] \log (1+u'D^*u/\zeta)\\
&=  \log \left( \Gamma \left( \dfrac{\zeta+q}{2}  \right)   \right)-\log \left( \Gamma \left( \dfrac{\zeta}{2}  \right)   \right)  - \dfrac{q}{2} \log(\nu \pi)\\
&- \dfrac{1}{2} \log |D^*|   -\left[ \dfrac{\zeta+q}{2} \right] \log \left(1+\sum_{t=1}^T \dfrac{u_t'u_t}{\zeta \nu_t^*}\right)
\end{align}

Rewritten in C, we need to pass in the first four terms \texttt{tconstants} as a double, \texttt{u} as an array of doubles, \texttt{zeta} as an int, \texttt{myq} as an array of ints, \texttt{meow} as an array of ints, \texttt{nut} as an array of doubles,  and \texttt{logft} as a double. The final calculation will be written into \texttt{logft}.

@@

&\leq \int \dfrac{f_\theta(y|u)^2 f_\theta(u)^2}{p_1 \grave{f}(u|0,D^*)} du\\

@@

First, we start with \eqref{eq:first}:
\begin{align}
\int \dfrac{ \left[ \nabla \log f_\theta (y|u) \right]^2 \left[ f_\theta(u,y)\right]^2}{\tilde{f}(u)} du 
&=\int \dfrac{ \left[ \nabla \log f_\theta (y|u) \right]^2 \left[ f_\theta(y|u)\right]^2 \left[ f_\theta(u)\right]^2}{\tilde{f}(u)} du \\
&\leq \int \dfrac{ \left[ \nabla \log f_\theta (y|u) \right]^2 \left[ f_\theta(y|u)\right]^2 \left[ f_\theta(u)\right]^2}{p_1\grave{f}(u|0,D^*)} du \\
&\propto \int \dfrac{ \left[ \nabla \log f_\theta (y|u) \right]^2 \left[ f_\theta(y|u)\right]^2 \exp(-u'D^{-1}u)}{p_1\left[ 1+u'D^*u/\zeta  \right]^{-(\zeta+q)/2}} du \\
&= \int \dfrac{ \left[ \nabla \log f_\theta (y|u) \right]^2 \left[ f_\theta(y|u)\right]^2 \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} }{p_1 \exp(u'D^{-1}u)} du \\
&= \int \dfrac{ \left[ \nabla \log f_\theta (y|u) \right]^2 \, \exp(2y'(X \beta+Zu)) \,  \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} }{ p_1 \, \exp(2 c(\eta)) \,\exp(u'D^{-1}u)} du \label{eq:soon}
\end{align}
By the Cauchy-Schwartz inequality, the integral in \eqref{eq:soon} exists if
\begin{align}
 & \int \dfrac{ \left[\dfrac{\partial}{\partial \beta} \log f_\theta (y|u) \right]^2 \, \exp(2y'(X \beta+Zu)) \,  \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} }{ p_1 \, \exp(2 c(\eta)) \,\exp(u'D^{-1}u)} du \nonumber \\
&\propto  \int \dfrac{ \left[X' c'(\eta) \right]^2 \, \exp(2y'(X \beta+Zu)) \,  \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} }{ p_1 \, \exp(2 c(\eta)) \,\exp(u'D^{-1}u)} du \label{eq:soon2}
\end{align}
exists, where $c'(\eta)$ is a vector with components $c'(\eta_i)$.  This will converge as long as no term in the numerator of the integrand grows more quickly than $\exp(u'D^{-1}u$. The second and third terms grow less quickly than $\exp(u'D^{-1}u$, so $\left[ c'(\eta)\right]^2$ is the only term to check.  If $y|u \sim$ Bernoulli, then 
\begin{align}
c'(\eta_i)= \dfrac{\exp(\eta_i)}{1+\exp(\eta_i)},
\end{align}
which does not grow more quickly than $\exp(u'D^{-1}u)$. If $y|u \sim$ Poisson, then
\begin{align}
c'(\eta_i)= e^{\eta_i},
\end{align}
which does not grow more quickly than $\exp(u'D^{-1}u)$. Therefore, \eqref{eq:first} exists and is finite, and 
\begin{align}
\var \left( \sum_{k=1}^m \nabla \log f_\theta (y|u_k) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) < \infty.
\end{align}
is finite.

Next, we move on to showing the existence of \eqref{eq:second}.
\begin{align}
\int \dfrac{ \left[ \nabla \log f_\theta (u) \right]^2 \left[ f_\theta(u,y)\right]^2}{\tilde{f}(u)} du &= 
\int \dfrac{ \left[ \nabla \log f_\theta (u) \right]^2 \left[ f_\theta(y|u)\right]^2  \left[ f_\theta(u)\right]^2}{\tilde{f}(u)} du\\
&\leq \int \dfrac{ \left[ \nabla \log f_\theta (u) \right]^2 \left[ f_\theta(y|u)\right]^2  \left[ f_\theta(u)\right]^2}{p_1 \grave{f}(u|0,D^*)} du
\end{align}
By the Cauchy-Schwartz inequality, the above integral exists as long as  
\begin{align}
\int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t} \log f_\theta (u) \right]^2 \left[ f_\theta(y|u)\right]^2  \left[ f_\theta(u)\right]^2}{p_1 \grave{f}(u|0,D^*)} du < \infty
\end{align}
for every $t=1,\ldots, T$. We see
\begin{align}
&\int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t} \log f_\theta (u) \right]^2 \left[ f_\theta(y|u)\right]^2  \left[ f_\theta(u)\right]^2}{p_1 \grave{f}(u|0,D^*)} du \\
&= \int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t} \log f_\theta (u) \right]^2 \exp(2y'(X\beta+Zu)   )  \left[ f_\theta(u)\right]^2}{p_1 \exp(2c(\eta))    \grave{f}(u|0,D^*)} du \\
&\propto \int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t} \log f_\theta (u) \right]^2 \exp(2y'(X\beta+Zu)   ) \,   \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} }{p_1 \exp(2c(\eta)) \, \exp(u'D^{-1}u)  } du\\
&= \int \dfrac{ \left[  -\dfrac{q_t}{2 \nu_t} + \dfrac{u_t'u_t}{2 \nu_t^2}   \right]^2 \exp(2y'(X\beta+Zu)   ) \,   \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} }{p_1 \exp(2c(\eta)) \, \exp(u'D^{-1}u)  } du
\end{align}
Because every term in the numerator of the integrand grows less quickly than $\exp(u'D^{-1}u)$, the integral exists.  Therefore, \eqref{eq:second} exists and 
\begin{align}
\var \left(\sum_{k=1}^m \nabla \log f_\theta (u_k) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) < \infty.
\end{align}

Lastly, we want to show the existence of \eqref{eq:third}. We see
\begin{align}
\int \dfrac{  \nabla \log f_\theta (u)\,  \nabla \log f_\theta (y|u) \, \left[ f_\theta(u,y)\right]^2}{\tilde{f}(u)} du &=
\int \dfrac{  \nabla \log f_\theta (u) \, \nabla \log f_\theta (y|u) \, \left[ f_\theta(y|u)\right]^2 \left[ f_\theta(u)\right]^2}{\tilde{f}(u)} du \\
&\leq
\int \dfrac{  \nabla \log f_\theta (u) \,  \nabla \log f_\theta (y|u)\,  \left[ f_\theta(y|u)\right]^2 \left[ f_\theta(u)\right]^2}{p_1 \grave{f}(u|0,D^*)} du \label{eq:this1}
\end{align}
Again, we use the Cauchy-Schwartz inequality to check the existence of \eqref{eq:this1}. If, for every $t=1,\ldots,T$,
\begin{align}
\int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t}  \log f_\theta (u) \right] \left[  \dfrac{\partial}{\partial \beta} \log f_\theta (y|u) \right]   \left[ f_\theta(y|u)\right]^2 \left[ f_\theta(u)\right]^2}{p_1 \grave{f}(u|0,D^*)} du \label{eq:this1}
\end{align}
exists, then \eqref{eq:this1} also exists. Continuing, we see
\begin{align}
& \int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t}  \log f_\theta (u) \right] \left[  \dfrac{\partial}{\partial \beta} \log f_\theta (y|u) \right]   \left[ f_\theta(y|u)\right]^2 \left[ f_\theta(u)\right]^2}{p_1 \grave{f}(u|0,D^*)} du\\
 &\propto \int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t}  \log f_\theta (u) \right] \left[  \dfrac{\partial}{\partial \beta} \log f_\theta (y|u) \right]   \left[ f_\theta(y|u)\right]^2 }{p_1 \, \exp(u'D^{-1}u) \, \grave{f}(u|0,D^*)} du\\
 &\propto \int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t}  \log f_\theta (u) \right] \left[  \dfrac{\partial}{\partial \beta} \log f_\theta (y|u) \right]   \left[ f_\theta(y|u)\right]^2 \, \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} }{p_1 \, \exp(u'D^{-1}u) } du\\
&\propto \int \dfrac{ \left[      -\dfrac{q_t}{2 \nu_t} + \dfrac{u_t'u_t}{2 \nu_t^2}        \right] \left[  X'(y-c'(\eta)) \right]   \left[ f_\theta(y|u)\right]^2 \, \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} }{p_1 \, \exp(u'D^{-1}u) } du\\
&= \int \dfrac{ \left[      -\dfrac{q_t}{2 \nu_t} + \dfrac{u_t'u_t}{2 \nu_t^2}        \right] \left[  X'(y-c'(\eta)) \right]   \, \exp(2y'(X\beta+Zu)) \, \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} }{p_1 \, \exp(u'D^{-1}u) \, \exp(2 c(\eta)) } du. \label{eq:this2}
\end{align}
We have already shown that $-c'(\eta)$ does not grow more quickly than $\exp(u'D^{-1}u)$. The other terms in the numerator of the integrand shown in \eqref{eq:this2} also grow more slowly than $\exp(u'D^{-1}u)$. We conclude that \eqref{eq:this2} exists, which implies that \eqref{eq:third} exists and
\begin{align}
\cov \left(\sum_{k=1}^m \nabla \log f_\theta (y|u_k) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \, , \, \sum_{k=1}^m \nabla \log f_\theta (u_k) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) < \infty.
\end{align}
This shows that
\begin{align}
\var \left( \sum_{k=1}^m     \nabla \log f_\theta(u_k,y)\dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) < \infty. 
\end{align}
Therefore, the gradient of the MCLA has finite variance and a central limit theorem.
